\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-21 paper/paper.tex
Updated Discussion

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA).  A PCA is justifiable in the limit where there is very high signal-to-noise ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

@@

We settled on a $K$ value of 6.  Figure \ref{fig:pcLfc} shows the first 9 principal components constructed using LFC lines.  There is clear structure in the first and second principal components.  \lz{The diagonal-banding structure seen in components 3, 5, and 6 we believe to be capturing detector fabrication defects. (Maybe, awaiting STA response)}  Components 3 and 4 show aberrant behavior on the edges of the two bluest echelle orders, where lower signal results in more variation in the line fits.  Later principal components become dominated by noise over coherent structure.

@@

\section{Discussion} \label{sec:discussion}
The endeavor to reach $0.1\,\mps$ precision necessitates improvements on all levels of spectroscopy, from the instrumentation to the data analysis.  In this paper, we present \name\, a hierarchical, non-parametric model for wavelength calibration.

Starting with a list of calibration lines with known wavelengths and well-fit line centers for each calibration exposure, \name\ will de-noise and interpolate the given lines into a full wavelength solution.  We show that \name\ returns a lower RMS than classic, parametric methods by over a factor of two.  Using \name\ wavelengths also reduced the RMS in a planet fit to derived RVs by \lz{some percent hopefully probs}.

\name\ levies the incredible stability of contemporary EPRV instruments and high density of lines made available by new calibration sources, such as LFCs and etalons, to achieve more accurate wavelengths.  Denser calibration lines allow us to move to more flexible, interpolated wavelength solutions.  Stabilized spectrograph hardware with few degrees of freedom allow us to use all calibration images in a given generation of stability to constrain the accessible calibration space of the spectrograph.  We have described only one, fairly simplistic implementation of \name\ here.  There are many options for both the de-noising and interpolation steps, as mentioned in Section \ref{sec:choices}.

An advantage of this implementation, where PCA is applied to line positions from all LFC exposures, is the ability to isolate exposures that exhibit errant variation, which is typically associated with flawed exposures.  This allowed us to quickly vet for problematic LFC exposures, which otherwise would have required visual inspection of all 1200+ LFC exposures.  In a classic framework where each calibration exposure is treated independently, these aberrant exposures would likely have persisted undetected and could potentially sway the resultant wavelength solutions.

On the other hand, the ability of PCA to pick up on any and all variation, regardless of source, also makes \name\ very sensitive to uncertainties in line fitting.  For example, we have seen that lower-signal lines that are harder to fit faithfully will have greater variety in returned line positions, which is in turn captured by the PCA.  High-fidelity line positions are essential to ensure the PCA is capturing variations in the spectrograph's calibration rather than changes in how well a line can be fit.

\name\ can be applied to any data that contains information about the calibration state of the spectrograph.  For example, though LFC and ThAr exposures are used as an example in this paper, \name\ would work exactly the same for an etalon or any other arc lamp with a list of lines and assigned wavelengths.

Incorporating calibration lines taken through a simultaneous fiber is trivial, especially if the simultaneous reference is also used while taking calibration images.  The calibration lines through the simultaneous fiber could simply be concatenated to the array of calibration lines run through the dimensionality reduction and de-noising step.  This not only allows the simultaneous calibration information to contribute to constructing the complete calibration space of the spectrograph, but also allows the simultaneous calibration information to uniquely pinpoint where the spectrograph is in that calibration space for any exposure, removing the need to interpolate the basis vector amplitudes.

Once we have defined a calibration space that captures all possible degrees of freedom for a stabilized spectrograph, then all exposures taken with the spectrograph contains information about where the spectrograph is located within that calibration space.  Theoretically, tellurics, lines in science exposures, or just the trace positions themselves could also be used to determine the calibration state of an instrument, negating the need for any additional wavelength-calibration source beyond defining the calibration space.

\lz{(Not sure where this belongs, honestly.)}  We caution that with any wavelength solution, there is a perfect degeneracy between what is defined as the ``center of the line'' and the resultant wavelength solution.  If, for example, a cross correlation method is used to extract RVs from the data, a systematic difference may be introduced depending on the definition of the center of the line.  In principle, the best way to mitigate this uncertainty would be to use a calibration source that looks like a star.

\name\ is designed and optimized for extreme-precision, radial-velocity spectrographs.  The errors returned by \name\ on \expres\ data renders the uncertainty from wavelength calibration negligible in the face of other noise sources, for example, stellar activity.  Though the focus was on \eprv\ instruments here, \name\ should be largely applicable to nearly any astronomical spectrograph.

>>>>

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA) \citep{wiki_pca}.  A PCA is justifiable \lz{(Hogg)} in the limit where there is very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

@@

We settled on a $K$ value of 6.  Figure \ref{fig:pcLfc} shows the first 9 principal components constructed using LFC lines.  There is clear structure in the first and second principal components.  \lz{The diagonal-banding structure seen in components 3, 5, and 6 we believe to be capturing detector fabrication defects. in \expres's detector or other optical elements}  Components 3 and 4 show aberrant behavior on the edges of the two bluest echelle orders, where lower signal results in more variation in the line fits.  Later principal components become dominated by noise over coherent structure.

@@

\section{Discussion} \label{sec:discussion}
We show that \name\ returns a lower RMS than classic, parametric methods by over a factor of two (Section \ref{sec:tests}).  Using \name\ wavelengths also reduced the RMS in a planet fit to derived RVs by \lz{some percent hopefully probs} (Section \ref{sec:test-rv}).  The endeavor to reach $0.1\,\mps$ radial-velocity precision necessitates improvements on all levels of spectroscopy, from the instrumentation to the data analysis.  We believe implementing \name, a hierarchical, non-parametric model for wavelength calibration, is a step towards reducing the error contribution from data reduction pipelines.

Starting with a list of calibration lines with known wavelengths and well-fit line centers for each calibration exposure, \name\ will de-noise and interpolate the given lines into a full wavelength solution.  \name\ levies the incredible stability of contemporary EPRV instruments and high density of lines made available by new calibration sources, such as LFCs and etalons, to achieve more accurate wavelengths.  \name\ therefore assumes dense enough calibration lines to properly constrain a non-parametric wavelength model, and that the instrument has low degrees of freedom (see Section \ref{sec:others} for further discussion).

Denser calibration lines allow us to move to more flexible wavelength models that can account for non-smooth features in the wavelength solution.  Stabilized spectrograph hardware insures the calibration state of the instrument is low-dimensional.  All calibration images in a given generation of stability can therefore be used to constrain the accessible calibration space of the spectrograph as well as where in that calibration space the spectrograph lies for each calibration exposures.  We have described only one, fairly simplistic implementation of \name\ here.  There are many options for both the de-noising and interpolation steps, as mentioned in Section \ref{sec:choices}.

An advantage of this implementation, where PCA is applied to line positions from all LFC exposures, is the ability to isolate exposures that exhibit errant variation, which is typically associated with flawed exposures.  This allowed us to quickly vet for problematic LFC exposures, which otherwise would have required visual inspection of all 1200+ LFC exposures.  In a classic framework where each calibration exposure is treated independently, these aberrant exposures would likely have persisted undetected and would likely have swayed the resultant wavelength solutions for at least an entire night.

On the other hand, PCA captures all variance, regardless of source.  Though \name\ only wants to capture variation in the instrument, the PCA is also sensitive to uncertainties and failures in the upstream line position fitting.  For example, we have seen that lower-signal lines that are harder to fit faithfully will have greater variety in returned line positions, which is in turn captured by the PCA.  In this sense, \name\ is actually a model of not just the instrument, but all the upstream choices used to drive live positions.  High-fidelity line positions are essential to ensure the PCA is capturing variations in just the spectrograph's calibration rather than changes in how well a line can be fit or other effects.

Along those lines, we caution that with any wavelength solution, there is a perfect degeneracy between what is defined as the ``position of the line'' and the resultant wavelength solution.  If, for example, a cross correlation method is used to extract RVs from the data, a systematic difference may be introduced depending on the definition of what exactly is the line position, whether it be the center of a Gaussian fit, the edge of a complicated PSF, etc.  In principle, the best way to mitigate this uncertainty would be to use a calibration source that looks like a star.

Compared to traditional methods, which involve fitting high-order polynomials, \name\ has several useful statistical properties.  \name\ is technically robust to localized issues with either the calibration source or the pipeline used to return line positions.  With an interpolated wavelength model, one errant line position will only effect the resultant wavelength model out to the closest, neighboring lines.  The effect of an outlier is diminished and kept local.  In contrast, a parametric fit will change from a single errant line, changing the wavelength solution for the entire echelle order or the entire exposure.  Through de-noising and outlier rejection, \name\ adds additional robustness against erroneous line positions.

The locality of the interpolation caries other benefits as well.  Manufacturing artifacts in the detector or other optical elements can lead to non-smooth structure in the wavelength solution that can not be captured by polynomials or other smooth functions (see \figurename~\ref{fig:resid2d}).  An interpolated model introduces greater flexibility, enabling the model to account for such high-order effects.  As discussed in Section \ref{sec:choice_wvp}, there are better and worse interpolators for the task, which may differ for different instruments and different calibration sources.  Instead of using an interpolator at all, there might be better results from implementing something more sophisticated, such as a kernel method or a Gaussian process with a kernel adapted for the specifics of an instrument.  There is in principle an enormous space of non-parametric methods to explore, which we leave outside the scope of this paper.

Similarly, PCA is just one of many possible dimensionality-reduction methods.  We chose to implement \name\ using PCA here for simplicity and computational tractability.  If \name\ was updated to a full probabilistic model, the PCA along with the interpolation model would have to be upgraded to something with better probabilistic properties.  Further discussion of other implementations of \name\ can be found in Section \ref{sec:choices}.

\name\ can be applied to any data that contains information about the calibration state of the spectrograph.  For example, though LFC and ThAr exposures are used as an example in this paper, \name\ would work similarly for an etalon or any other arc lamp with a list of lines and assigned wavelengths.  Modifications would have to be made with an etalon to include the free spectral range (FSR) as a free parameter that is fit for along with the low-dimensional basis.

Incorporating calibration lines taken through a simultaneous fiber would also be fairly simple, especially if the simultaneous reference is also used while taking calibration images.  The calibration lines through the simultaneous fiber could simply be concatenated to the array of calibration lines when the low-dimensional basis is constructed.  This not only allows the simultaneous calibration information to contribute to constructing the complete calibration space of the spectrograph, but also allows the simultaneous calibration information to uniquely pinpoint where the spectrograph is in that calibration space for any exposure, removing the need to interpolate the basis vector amplitudes from calibration exposures to science exposures.

Once we have defined a calibration space that captures all possible degrees of freedom for a stabilized spectrograph, there are many options for pinpointing where the spectrograph is located within that calibration space.  Good housekeeping data, such as temperature or pressure could be used in addition to or instead of time (as mentioned in Section \ref{sec:choice_avt}).  Telemetry that is seen to be correlated with the calibration state of the spectrograph can even be added to the data used to construct the low-dimensional basis.  Furthermore, all exposures taken with the spectrograph in principle contains information about the calibration state of the spectrograph.  Theoretically, tellurics, lines in science exposures, or just the trace positions themselves could also be used to determine an instrument's calibration state, negating the need for any additional wavelength-calibration source beyond defining the calibration space.

\name\ is designed and optimized for extreme-precision, radial-velocity spectrographs.  In the battle to construct a high-fidelity data pipeline for extreme-precision radial-velocity measurements, we believe \name\ represents a step towards mitigating the error from wavelength calibration, as demonstrated by tests using \expres\ data (Section \ref{sec:tests}).  Though the focus was on \eprv\ instruments here, \name\ should be largely applicable to nearly any astronomical spectrograph.
